### 2022-03-11 bools/bools.tex
more

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the higher-order provers
Leo-III \cite{sb-21-leo3} and Satallax \cite{cb-12-satallax}, such as
elimination of Leibniz equality, primitive instantiation, and treatment of choice
operator \cite{pa-01-classical-ty-thy}. We have also designed new rules that use
higher-order unification to resolve Boolean formulas that are hoisted to literal
level, delay clausification of nonatomic literals, reason about formulas under
$\lambda$-binders, and many others. Even though the rules we use are inspired by
the ones of refutationally complete higher-order provers, we do not guarantee
completeness of our extension of \lsup{}. Using the insights that we gained with
this incomplete extension of \lsup{} to full higher-order logic, Bentkamp
designed a complete calculus for full higher-order logic---\osup{}
\cite{bbtv-21-full-ho-sup}.

@@

We base our work (and parts of the following text) on Bentkamp et al.'s
\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. We
extend the syntax of this logic by adding logical connectives to the signature.
The semantic of the logic is extended by interpreting Boolean type $o$ as a
two-element domain. This amounts to extending the logic to full higher-order
logic. Taking a different perspective, this logic is an extension of the one
presented in Chapter \ref{ch:ehoh} with $\lambda$-abstraction, polymorphic
types, and native logical connectives. For reference, we provide a
definition of this polymorphic logic.

@@

\looseness=-1
Given a formula $f$ we call its Boolean subterm $f|_p$ a \emph{top-level
Boolean} if for all proper prefixes $q$ of $p$, the head of $f|_q$ is a logical
constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the
formula $f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q}) \ior
\inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while $f|_{1.2.1}$ is a
nested Boolean, as well as its subterms. First-order logic allows only top-level
Booleans, whereas nested Booleans are characteristic for higher-order logic. 
Note that while our calculus works with clausal representation of the problem,
its higher-order nature allows representing formula $f$ as a singleton clause
containing only literal $f$.

@@

\looseness=-1
The axiom of choice is notoriously difficult to handle efficiently in higher-order provers. Andrews formulates
this axiom as $\iforall p.\, (\iexists
x.\, p \, x) \iimplies p \, (\varepsilon \, p)$, where $\varepsilon :
\forallty{\alpha}(\alpha \rightarrow o) \rightarrow \alpha$ denotes the \emph{choice
operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
\, (\varepsilon \, P) \eq \itrue$. Since the term $P \, X$ matches any Boolean
term in the proof state, this axiom is very explosive. Therefore, Leo-III
\cite{sb-21-leo3} recognizes definitions of all symbols that correspond to choice and deals with
them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
added to set of choice functions $\mathit{CF}$ (which initially contains just
$\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
instantiate the axiom of choice. We reused the method of recognizing choice
functions, but generalized the rule for creating the instance of
the axiom (assuming $\xi \in \mathit{CF}$ and $X$ and $z$ are fresh variables):
%
$$\namedinference{Choice}
               {C[\xi \, t]}
               { X \, (t \, Y) \noteq \itrue \llor X \, (t \, (\xi \, (\lam{z}{X \, (t \, z)})) ) \eq \itrue }$$
%
Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
arbitrary context around $t$, the chosen instantiation for $P$ from axiom of choice;
the variable $X$ can later be replaced by imitations of logical symbols to create more
complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, depending on
Zipperposition options, $D$ will either be deleted or kept. Note that $D$
will not subsume its instances, since the matching algorithm Zipperposition uses is
too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
\looseness=-1

Most provers natively support both functional and Boolean extensionality reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify
first-order unification to return unification constraints consisting of pairs of
terms of functional type, whereas Steen relies on the unification rules of
Leo-III's calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality.
As a pragmatic extension of the lambda-superposition calculus, Bentkamp et al. \cite{bbtvw-21-sup-lam} propose to alter core generating inference rules of \lsup{}
to support functional extensionality. Instead of requiring that terms
involved in the inference are unifiable, it is required that they can be
decomposed into \emph{disagreement pairs} such that at least one of the
disagreement pairs is of functional type. Disagreement pairs of terms $s$ and
$t$ of the same type are defined inductively using function $\textsf{dp}$:
$\textsf{dp}(s,t) = \emptyset$ if $s$ and $t$ are equal; $\textsf{dp}(a \,
\tuplen{s}, b \, \tuple{t}{m}) = \{(a \,
\tuplen{s}, b \, \tuple{t}{m})\}$ if $a$ and $b$ are different heads;
$\textsf{dp}(\lam{x}{s}, \lam{y}{t}) = \{ (\lam{x}{s}, \lam{y}{t}) \}$;
$\textsf{dp}(a \, \tuplen{s}, a \, \tuplen{t}) = \bigcup_{i=1}^{n}
\textsf{dp}(s_i, t_i)$. Then the extensionality rules are stated as follows:
\pagebreak[2]
\begin{align*}
  & \namedinference{AbsSup}
  {s \eq t \llor C \qquad u[s'] \mathrel{\dot{\eq}} v \llor D}
  {\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor  u[t] \mathrel{\dot{\eq}} v \llor C \llor D}} \\
  &\namedinference{AbsER}{s \noteq s' \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \cdots \llor  C}} \\
  &\namedinference{AbsEF}{s \eq t \llor  s' \eq u \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor t \noteq u \llor  s' \eq u \llor C}}
\end{align*}

@@

\begin{sloppypar}
  Zipperposition restricts \infname{AbsSup} by
  requiring that $s$ and $s'$ are not of function or Boolean types. If the terms are of function type, our experience is
  that better treatment of function extensionality is to apply fresh free
  variables (or Skolem terms, depending on the sign
  \cite{bbtvw-21-sup-lam}) to both sides of a (dis)equation to reduce it to
  a first-order literal; Boolean extensionality is usually better supported by
  applying \infname{IC} on the top-level Boolean term. Thus, for the following
  discussion we can assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
  and $s'$ have the same head, and a functional or Boolean subterm. To speed up
  retrieval of such terms, we added an index that maps symbols to positions in
  clauses where they appear as a head of a term that has a functional or Boolean
  subterm. This index will be empty for first-order problems, incurring no
  overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply the \infname{Abs} rules if all
  disagreement pairs have at least one side whose head is a variable; those will
  be dealt with more efficiently using the core
  rules of the superposition calculus. To simplify the resulting clauses of \infname{Abs} rules,
  we also eagerly remove literals $s_i \not\eq s'_i$ by applying the first
  unifier returned by terminating, pragmatic variant of unification algorithm described in 
  Sect.~\ref{sec:unif:the-unification-procedure}.
\end{sloppypar}

@@

The winner of the higher-order theorem division of the 2019 edition of CASC \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
crucial advantage over Zipperposition: it is based on higher-order
tableaux, and as such it does not require formulas to be converted to clauses. The advantage of tableaux is that once it instantiates a variable with a
term, this instantiation naturally propagates through the whole formula. In
Zipperposition, which is based on \lsup{}, the original formula
is clausified and instantiating a variable in a clause $C$ does not automatically instantiate
it in all clauses that are results of clausification of the same formula as $C$.
To mitigate this issue, we have created extensions of equality resolution and equality factoring
that take Boolean extensionality into account:
%
\begin{align*}
  & \namedinference{BoolER}
  { s \eq s' \llor C}
  {\sigmacl{C}}
  &&
  \namedinference{BoolEF$+-$}
  {P\,\tuplen{s} \eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}} \\
  & \namedinference{BoolEF$-+$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \eq \itrue \llor C}
  {\sigmacl{s' \eq \itrue \llor C}} &&
  \namedinference{BoolEF$-$$-$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}}
\end{align*}
%
All side conditions except for the ones concerning the unifiability of terms are
as in the original equality resolution and equality factoring rules. In rule
\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \inot s' $. In the $+-$
and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ P \, \tuplen{s}
$ and $\inot s'$, and in the remaining version it unifies $ P \, \tuplen{s} $ and
$s'$. Intuitively, these rules bring Boolean (dis)equations in the appropriate
form for application of the corresponding base rules. In particular,
for \infname{BoolER} we interpret the equation $s \eq s'$ as $s \not\eq \inot s'$ which
allows us to simulate \infname{ER} inference from Sect.~\ref{sec:pre:rules}. 
For \infname{BoolEF} family of rules, we convert disequations into equations
using the trick of negating one side to simulate \infname{EF}.
% It suffices to consider 
% literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
% $s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
% rewriting into $ s \not\eq t $).
The rule \infname{BoolER} only considers literals of the form $s \eq t$. 
Equivalences $s \iequiv t = \itrue$ and disequivalences $(s \iequiv t) \not\eq \itrue$ are automatically simplified to $s \eq t$ or $s \not\eq t$.
The example  \verb|SET557^1| in Sect.~\ref{sect:bool:examples} illustrates how 
\infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.

@@

\vspace{2\jot}
\begin{tabular}{ccc}
  $\namedsimp{DC$_{\iand}$}
  { (f \iand g) \eq \itrue \llor C}
  { f \eq \itrue \llor C \quad g \eq \itrue \llor C }$ &
  $\namedsimp{DC$_{\ior}$}
  { (f \ior g) \eq \itrue \llor C}
  { f \eq \itrue \llor g \eq \itrue  \llor C }$ &
  $\namedsimp{DC$_{\iimplies}$}
  { (f \iimplies g) \eq \itrue \llor C}
  { f \not\eq \itrue \llor g \eq \itrue  \llor C } $ \\[2\jot]
  $\namedsimp{DC$_{\inot}$}
  { (\inot f) \eq \itrue \llor C}
  { f \not\eq \itrue \llor C}$ &
  $\namedsimp{DC$_{\iforall}$}
  { (\iforall x. \, f) \eq \itrue \llor C}
  { \{x \mapsto Y\}(f) \llor C }$ &
  $\namedsimp{DC$_{\iexists}$}
  { (\iexists x.\, f) \eq \itrue \llor C}
  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}(f) \llor C }$ 
  \\[2\jot]
  \multicolumn{3}{c}{$\namedsimp{DC$_\eq$}
  { f \eq g \llor C}
  { f \not\eq \itrue \lor g\eq\itrue \llor C \quad f \eq \itrue \lor g\not\eq\itrue \llor C}$}
\end{tabular}
\vspace{2\jot}

@@

Leo-III and Satallax are cooperative theorem provers---they periodically invoke
first-order provers to finish the proof attempt. Leo-III uses CVC4, E and
iProver \cite{kk-08-iprover} as backends, and Satallax uses Ehoh
% \cite{EHOH CHAPTER}
 as backend. Zipperposition can use Ehoh as backend as well. To test how successful each calculus is, we run
the cooperative provers in two versions: \emph{pure}, which disables backends,
and \emph{coop} which uses all supported backends.

\begin{figure}[t]
  \label{fig:higher-order-provers}
  \begin{center}
    \def\arraystretch{1.1}%
    \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c}
      & CVC4             & Leo-III & Satallax              & Vampire & Zipperposition
      \MIDLINE
      pure    & 1806 (5) &  1627 (0) & 2067 (0)            & 1924 (7) & 1980 (\phantom{0}0)                \\
      coop    & --       &  2085 (3) & \textbf{2214 (9)}   & --       & 2190 (17)
    \end{tabular}}
    \caption{Comparison with other higher-order provers}
  \end{center}
\end{figure}

\looseness=-1
In both pure and cooperative mode, Satallax is the winner. Zipperposition comes
in close second, showing that our approach is a promising basis for further
extensions. Indeed, as we show in Chapter \ref{ch:ho-techniques}, with
fine-tuning of the heuristics and further extensions to the calculus, this
approach can outperform all other competitive provers with a large margin.

\subsection{Discussion}
\label{sect:bool:discussion}

>>>>

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the higher-order provers
Leo-III \cite{sb-21-leo3} and Satallax \cite{cb-12-satallax}, such as
elimination of Leibniz equality, primitive instantiation, and treatment of choice
operator \cite{pa-01-classical-ty-thy}. We have also designed new rules that use
higher-order unification to resolve Boolean formulas that are hoisted to literal
level, delay clausification of nonatomic literals, reason about formulas under
$\lambda$-binders, and many others. Even though the rules we use are inspired by
the ones of refutationally complete higher-order provers, we do not guarantee
completeness of our extension of \lsup{}. Using the insights that we gained with
this incomplete extension of \lsup{} to full higher-order logic, Bentkamp et al.
designed a complete calculus for full higher-order logic---\osup{}
\cite{bbtv-21-full-ho-sup}.

@@

We base our work (and parts of the following text) on Bentkamp et al.'s
\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. We
extend the syntax of this logic by adding logical connectives to the signature.
The semantic of the logic is extended by interpreting Boolean type $o$ as a
two-element domain. This amounts to extending the logic to full higher-order
logic. Taking a different perspective, this logic is an extension of the one
presented in Chapter \ref{ch:ehoh} with $\lambda$-abstraction, polymorphic
types, and native logical connectives. For reference, we provide a
description of this polymorphic logic. For the notions that are not defined here,
we assume the definitions introduced in Sect.~\ref{ch:pre}.

@@

\looseness=-1
Given a formula $f$ we call its Boolean subterm $f|_p$ a \emph{top-level
Boolean} if for all proper prefixes $q$ of $p$, the head of $f|_q$ is a logical
constant. Otherwise, we call it a \emph{nested Boolean}. For example, in the
formula $f = \cst{h} \, \cst{a} \ieq \cst{g} \, (\cst{p} \iimplies \cst{q}) \ior
\inot\cst{p}$, $f|_1$ and $f|_2$ are top-level Booleans, while $f|_{1.2.1}$ is a
nested Boolean, as well as its subterms. First-order logic allows only top-level
Booleans, whereas nested Booleans are characteristic for higher-order logic.

% Note that while our calculus works with clausal representation of the problem,
% its higher-order nature allows representing formula $f$ as a singleton clause
% containing only literal $f$.
Our calculus works with the clausal structure described in Sect.~\ref{sec:pre:clauses}.
However, its higher-order nature allows representing formula $f$ as a singleton clause
containing only literal $f$.

@@

\looseness=-1
The axiom of choice is notoriously difficult to handle efficiently in higher-order provers. Andrews formulates
this axiom as $\iforall p.\, (\iexists
x.\, p \, x) \iimplies p \, (\varepsilon \, p)$, where $\varepsilon :
\forallty{\alpha}(\alpha \rightarrow o) \rightarrow \alpha$ denotes the \emph{choice
operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
\, (\varepsilon \, P) \eq \itrue$. Since the term $P \, X$ matches any Boolean
term in the proof state, this axiom is very explosive. Therefore, Leo-III
\cite{sb-21-leo3} heuristically recognizes symbols that correspond to choice and deals with
them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
added to set of choice functions $\mathit{CF}$ (which initially contains just
$\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
instantiate the axiom of choice. We reused the method of recognizing choice
functions, but generalized the rule for creating the instance of
the axiom (assuming $\xi \in \mathit{CF}$ and $X$ and $z$ are fresh variables):
%
$$\namedinference{Choice}
               {C[\xi \, t]}
               { X \, (t \, Y) \noteq \itrue \llor X \, (t \, (\xi \, (\lam{z}{X \, (t \, z)})) ) \eq \itrue }$$
%
Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
arbitrary context around $t$, the chosen instantiation for $P$ from axiom of choice;
the variable $X$ can later be replaced by imitations of logical symbols to create more
complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, depending on
Zipperposition options, $D$ will either be deleted or kept. Note that $D$
will not subsume its instances, since the matching algorithm Zipperposition uses is
too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
\looseness=-1

Most provers natively support both functional and Boolean extensionality reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify
first-order unification to return unification constraints consisting of pairs of
terms of functional type, whereas Steen relies on the unification rules of
Leo-III's calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality.
As a pragmatic extension of the \lsup{} calculus, Bentkamp et al. \cite{bbtvw-21-sup-lam} propose to alter core generating inference rules of \lsup{}
to support functional extensionality. Instead of requiring that terms
involved in the inference are unifiable, it is required that they can be
decomposed into \emph{disagreement pairs} such that at least one of the
disagreement pairs is of functional type. Disagreement pairs of terms $s$ and
$t$ of the same type are defined inductively using function $\textsf{dp}$:
$\textsf{dp}(s,t) = \emptyset$ if $s$ and $t$ are equal; $\textsf{dp}(a \,
\tuplen{s}, b \, \tuple{t}{m}) = \{(a \,
\tuplen{s}, b \, \tuple{t}{m})\}$ if $a$ and $b$ are different heads;
$\textsf{dp}(\lam{x}{s}, \lam{y}{t}) = \{ (\lam{x}{s}, \lam{y}{t}) \}$;
$\textsf{dp}(a \, \tuplen{s}, a \, \tuplen{t}) = \bigcup_{i=1}^{n}
\textsf{dp}(s_i, t_i)$. Then the extensionality rules are stated as follows:
\pagebreak[2]
\begin{align*}
  & \namedinference{AbsSup}
  {s \eq t \llor C \qquad u[s'] \mathrel{\dot{\eq}} v \llor D}
  {\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor  u[t] \mathrel{\dot{\eq}} v \llor C \llor D}} \\
  &\namedinference{AbsER}{s \noteq s' \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \cdots \llor  C}} \\
  &\namedinference{AbsEF}{s \eq t \llor  s' \eq u \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor t \noteq u \llor  s' \eq u \llor C}}
\end{align*}

@@

\begin{sloppypar}
  Zipperposition restricts \infname{AbsSup} by
  requiring that $s$ and $s'$ are not of function or Boolean types. If the terms are of function type, our experience is
  that better treatment of function extensionality is to apply fresh free
  variables (or Skolem terms, depending on the sign
  \cite{bbtvw-21-sup-lam}) to both sides of a (dis)equation to reduce it to
  a first-order literal; Boolean extensionality is usually better supported by
  applying \infname{IC} on the top-level Boolean term. Thus, for the following
  discussion we can assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
  and $s'$ have the same head, and a functional or Boolean subterm. To speed up
  retrieval of such terms, we added an index that maps symbols to positions in
  clauses where they appear as a head of a term that has a functional or Boolean
  subterm. This index will be empty for first-order problems, incurring no
  overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply the \infname{Abs} rules if all
  disagreement pairs have at least one side whose head is a variable; those will
  be dealt with more efficiently using the core rules of the superposition
  calculus. To simplify the resulting clauses of \infname{Abs} rules, we also
  eagerly remove literals $s_i \not\eq s'_i$ using a simplifying version of
  \infname{ER} (Sect.~\ref{sec:pre:rules}). In particular, we apply the first
  unifier returned by the terminating, pragmatic variant of unification
  algorithm described in Sect.~\ref{sec:unif:the-unification-procedure}, and
  remove the premise after conclusion of \infname{ER} is computed.
\end{sloppypar}

@@

The winner of the higher-order theorem division of the 2019 edition of CASC \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
crucial advantage over Zipperposition: it is based on higher-order
tableaux, and as such it does not require formulas to be converted to clauses. The advantage of tableaux is that once it instantiates a variable with a
term, this instantiation naturally propagates through the whole formula. In
Zipperposition, which is based on \lsup{}, the original formula
is clausified and instantiating a variable in a clause $C$ does not automatically instantiate
it in all clauses that are results of clausification of the same formula as $C$.
To mitigate this issue, we have created extensions of equality resolution and equality factoring
that take Boolean extensionality into account:
%
\begin{align*}
  & \namedinference{BoolER}
  { s \eq s' \llor C}
  {\sigmacl{C}}
  &&
  \namedinference{BoolEF$+-$}
  {P\,\tuplen{s} \eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}} \\
  & \namedinference{BoolEF$-+$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \eq \itrue \llor C}
  {\sigmacl{s' \eq \itrue \llor C}} &&
  \namedinference{BoolEF$-$$-$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}}
\end{align*}
%
All side conditions except for the ones concerning the unifiability of terms are
as in the original equality resolution and equality factoring rules. In rule
\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \inot s' $. In the $+-$
and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ P \, \tuplen{s}
$ and $\inot s'$, and in the remaining version it unifies $ P \, \tuplen{s} $ and
$s'$. Intuitively, these rules bring Boolean (dis)equations in the appropriate
form for application of the corresponding base rules. In particular,
for \infname{BoolER} we interpret the equation $s \eq s'$ as $s \not\eq \inot s'$ which
allows us to simulate \infname{ER} inference from Sect.~\ref{sec:pre:rules}. 
For the \infname{BoolEF} family of rules, we convert disequations into equations
using the trick of negating one side to simulate \infname{EF}.
% It suffices to consider 
% literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
% $s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
% rewriting into $ s \not\eq t $).
The rule \infname{BoolER} only considers literals of the form $s \eq t$. 
Equivalences $s \iequiv t = \itrue$ and disequivalences $(s \iequiv t) \not\eq \itrue$ are automatically simplified to $s \eq t$ or $s \not\eq t$.
The example  \verb|SET557^1| in Sect.~\ref{sect:bool:examples} illustrates how 
\infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.

@@

\vspace{2\jot}
\begin{tabular}{ccc}
  $\namedsimp{DC$_{\iand}$}
  { (g \iand h) \eq \itrue \llor C}
  { g \eq \itrue \llor C \quad h \eq \itrue \llor C }$ &
  $\namedsimp{DC$_{\ior}$}
  { (g \ior h) \eq \itrue \llor C}
  { g \eq \itrue \llor h \eq \itrue  \llor C }$ &
  $\namedsimp{DC$_{\iimplies}$}
  { (g\iimplies h) \eq \itrue \llor C}
  { g\not\eq \itrue \llor h \eq \itrue  \llor C } $ \\[2\jot]
  $\namedsimp{DC$_{\inot}$}
  { (\inot g) \eq \itrue \llor C}
  { g\not\eq \itrue \llor C}$ &
  $\namedsimp{DC$_{\iforall}$}
  { (\iforall x. \, g) \eq \itrue \llor C}
  { \{x \mapsto Y\}(g) \llor C }$ &
  $\namedsimp{DC$_{\iexists}$}
  { (\iexists x.\, g) \eq \itrue \llor C}
  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}(g) \llor C }$ 
  \\[2\jot]
  \multicolumn{3}{c}{$\namedsimp{DC$_\eq$}
  { g\eq h \llor C}
  { g\not\eq \itrue \lor h\eq\itrue \llor C \quad g \eq \itrue \lor h\not\eq\itrue \llor C}$}
\end{tabular}
\vspace{2\jot}

@@

Leo-III and Satallax are cooperative theorem provers---they periodically invoke
first-order provers to finish the proof attempt. Leo-III uses CVC4, E and
iProver \cite{kk-08-iprover} as backends, and Satallax uses Ehoh
% \cite{EHOH CHAPTER}
 as backend. Zipperposition can use Ehoh as backend as well. To test how successful each calculus is, we run
the cooperative provers in two versions: \emph{uncoop}, which disables backends,
and \emph{coop} which uses all supported backends.

\begin{figure}[t]
  \label{fig:higher-order-provers}
  \begin{center}
    \def\arraystretch{1.1}%
    \relax{\begin{tabular}{@{}l@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c@{\hskip 2em}c}
      & CVC4             & Leo-III & Satallax              & Vampire & Zipperposition
      \MIDLINE
      uncoop    & 1806 (5) &  1627 (0) & 2067 (0)            & 1924 (7) & 1980 (\phantom{0}0)                \\
      coop    & --       &  2085 (3) & \textbf{2214 (9)}   & --       & 2190 (17)
    \end{tabular}}
    \caption{Comparison with other higher-order provers}
  \end{center}
\end{figure}

\looseness=-1
In both uncooperative and cooperative mode, Satallax is the winner. Zipperposition comes
in close second, showing that our approach is a promising basis for further
extensions. Indeed, as we show in Chapter \ref{ch:ho-techniques}, with
fine-tuning of the heuristics and further extensions to the calculus, this
approach can outperform all other competitive provers with a large margin.

\section{Discussion}
\label{sect:bool:discussion}

### 2022-03-11 pre/pre.tex
more

\begin{enumerate}
    \item Bentkamp, Blanchette, Cruanes, and Waldmann designed a complete
    superposition calculus for higher-order logic devoid of
    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. We call
    this calculus \lfsup{}.
    \item Alexander Bentkamp, Blanchette, Tourret,  Vukmirović, and Waldmann
    extended this calculus to support $\lambda$-abstraction
    \cite{bbtvw-21-sup-lam}. We call it \lsup{}.
    \item Bentkamp, Blanchette, Tourret, and Vukmirovi\'c extended \lsup{} with
    support for first-class Booleans \cite{bbtv-21-full-ho-sup}. This amounts to
    designing the complete calculus for full higher-order logic. We call this
    calculus \osup{}.
\end{enumerate}

>>>>

\begin{enumerate}
    \item Bentkamp, Blanchette, Cruanes, and Waldmann designed a complete
    superposition calculus for higher-order logic devoid of
    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. This calculus is called \emph{Boolean-free $\lambda$-free higher-order superposition calculus} 
    and we shortly refer to it as \lfsup{}.
    \item \begin{sloppypar}
    Bentkamp, Blanchette, Tourret,  Vukmirović, and Waldmann
    extended this calculus to support $\lambda$-abstraction
    \cite{bbtvw-21-sup-lam}. They call this new calculus \emph{Boolean-free $\lambda$-superposition calculus},
    and we shortly refer to it as \lsup{}.
    \end{sloppypar}        
    \item Bentkamp, Blanchette, Tourret, and Vukmirovi\'c extended \lsup{} with
    support for first-class Booleans \cite{bbtv-21-full-ho-sup}. This amounts to
    designing the complete calculus for full higher-order logic. This
    calculus is called Boolean $\lambda$-superposition, or shortly \osup{}.
\end{enumerate}
%
Note that we use the name $\lambda$-superposition interchangeably for \lsup{}
and \osup{} where the context gives enough information to disambiguate this name.
